---
title: "Assignment_3_cwood19"
author: "Chris Wood"
date: "2026-06-09"
output: pdf_document
---

0: Partition the data into training and validation sets 60/40
Fields: Age, Experience, Income, Family, CCAvg, Mortgage, Securities.Account, CD.Account, Online, CreditCard, Education, Personal.Loan

```{r}
# Load packages
library(caret)
library(fastDummies)

# Load dataset 
UniversalBank <- read.csv("UniversalBank.csv")

# Remove unused variables
UniversalBank <- UniversalBank[, !(names(UniversalBank) %in% c("ID", "ZIP.Code"))]

# Handle extra Education variables with dummies
UniversalBank <- dummy_cols(
  UniversalBank,
  select_columns = "Education",
  remove_first_dummy = FALSE,
  remove_selected_columns = TRUE
)

# Set reproducibility
set.seed(123)

# Train the model for 60% / p = .6; y = target variable
train_index <- createDataPartition(
  UniversalBank$Personal.Loan,
  p = 0.60,
  list = FALSE
)

train_data <- UniversalBank[train_index, ]
valid_data <- UniversalBank[-train_index, ]

```

A. Create a pivot table for the training data with Online as a column variable, CreditCard as a row variable, and Personal.Loan as a secondary row variable. The values inside the table should convey the count. In R use functions melt() and cast(), or function table().

```{r}

# Generate pivot table with table()
pivot_table <- table(
  CreditCard = train_data$CreditCard,
  Personal.Loan = train_data$Personal.Loan,
  Online = train_data$Online
)

ftable(pivot_table)

```

A) The table identifies that more people are using online banking than not, and those that use online banking more frequently accept loans.

B. Consider the task of classifying a customer who owns a bank credit card and is actively using online banking services. Looking at the pivot table, what is the probability that this customer will accept the Personal.Loan offer? [This is the probability of Personal.Loan acceptance (Personal.Loan = 1) conditional on having a bank credit card (CreditCard = 1) and being an active user of online banking services (Online = 1)].

```{r}

# P(Personal.Loan = 1 / CreditCard = 1, Online = 1)
loan_yes <- pivot_table["1", "1", "1"]

total_cc_online <- sum(pivot_table["1", , "1"])

prob_pivot <- loan_yes / total_cc_online

prob_pivot

```

B) Based on the condition of having a CC and being an online service user, there's an estimated 10.7% chance of accepting a loan.

C. Create two separate pivot tables for the training data. One will have Personal.Loan (rows) as a function of Online (columns) and the other will have Personal.Loan (rows) as a function of CreditCard.

```{r}
loan_online <- table(
  Personal.Loan = train_data$Personal.Loan,
  Online = train_data$Online
)

loan_online

loan_cc <- table(
  Personal.Loan = train_data$Personal.Loan,
  CreditCard = train_data$CreditCard
)

loan_cc
```

D. Compute the following quantities [P(A | B) means “the probability of A given B”]:
1. P(CreditCard = 1 | Personal.Loan = 1) (the proportion of credit card holders among the loan acceptors)
2. P(Online = 1 | Personal.Loan = 1)
3. P(Personal.Loan = 1) (the proportion of loan acceptors)
4. P(CreditCard = 1 | Personal.Loan = 0)
5. P(Online = 1 | Personal.Loan = 0)
6. P(Personal.Loan = 0)

```{r}

### D.1 ###
p_cc_given_loan1 <-
  loan_cc["1","1"] / sum(loan_cc["1",])

p_cc_given_loan1

### D.2 ###
p_online_given_loan1 <-
  loan_online["1","1"] / sum(loan_online["1",])

p_online_given_loan1

### D.3 ###
p_loan1 <-
  sum(train_data$Personal.Loan == 1) /
  nrow(train_data)

p_loan1

### D.4 ###
p_cc_given_loan0 <-
  loan_cc["0","1"] / sum(loan_cc["0",])

p_cc_given_loan0

### D.5 ###
p_online_given_loan0 <-
  loan_online["0","1"] / sum(loan_online["0",])

p_online_given_loan0

### D.6 ###
p_loan0 <-
  sum(train_data$Personal.Loan == 0) /
  nrow(train_data)

p_loan0

```

E. Use the quantities computed above to compute the naive Bayes probability P(Personal.Loan = 1 | CreditCard = 1, Online = 1).

```{r}

# Calculate numerator: P(CC = 1 | L = 1) P(O = 1 | L = 1) P(L = 1)
numerator <-
  p_cc_given_loan1 *
  p_online_given_loan1 *
  p_loan1

# Calculate denominator: P(CC = 1 | L = 1) P(O = 1 | L = 1) P(L = 1) + P(CC = 1 | L = 0) P(O = 1 | L = 0) P(L = 0)
denominator <-
  numerator +
  p_cc_given_loan0 *
  p_online_given_loan0 *
  p_loan0

# Calculate P(L = 1 | CC = 1, O = 1) = numerator / denominator
nb_prob <- numerator / denominator

nb_prob

```

NB probability computed at 0.1105637 with current probability P(Loan = 1 | CC = 1, Online = 1)

F. Compare this value with the one obtained from the pivot table in (B). Which is a more accurate estimate?

```{r}

# Generate values
prob_pivot
nb_prob

```

F) Difference between prob_pivot & nb_prob = .0034208 or .34%, which is relatively insignificant, but the pivot table is closer to 1 and computes directly from the observed data rather than NB making assumptions about conditional independence between CC & Online given Loan.

G. Which of the entries in this table are needed for computing P(Personal.Loan = 1 | CreditCard = 1, Online = 1)? Run naive Bayes on the data. Examine the model output on training data, and find the entry that corresponds to P(Personal.Loan = 1 | CreditCard = 1, Online = 1). Compare this to the number you obtained in (E)

```{r}

# Add library for naiveBayes function
library(e1071)

# Fit the model
nb_model <- naiveBayes(
  as.factor(Personal.Loan) ~ CreditCard + Online,
  data = train_data
)

nb_model

# Display the model entries
nb_model$tables$CreditCard
nb_model$tables$Online

```

NB model uses the prior probabilities P(Personal.Loan = 1) = .0927 and P(Personal.Loan = 0) = .9073 with the conditional probabilities (.3273, .6439, .2910, and .5952) to yield the results forP(Personal.Loan = 1 | CreditCard = 1, Online = 1) = .1106, which very closely aligns to the pivot table's evaluation at .1071. 