---
title: "Final"
author: "Chris Wood"
date: "2026-06-23"
output: pdf_document
---

Clean / prepare data for relevant modeling
```{r, results = 'hide'}

# Load packages
library(tidyverse)
library(caret)
library(cluster)

set.seed(4827)

# Load dataset
fuel <- read_csv("monthly_fuel_costs.csv")

# Data overview
#glimpse(fuel)

# Null percentages
sapply(
  fuel[c("mercury_content_ppm",
         "moisture_content_pct",
         "chlorine_content_ppm")],
  function(x) mean(x == "null" | is.na(x)) * 100
)

# Uniqueness testing
sapply(fuel, function(x) length(unique(x)))

#table(fuel$data_maturity)
#table(fuel$fuel_cost_per_mmbtu_source)
#table(fuel$fuel_type_code_pudl)

# Data cleaning and selecting relevant fields for fuel quality modeling
fuel_clean <- fuel %>%
  select(
    fuel_mmbtu_per_unit,
    sulfur_content_pct,
    ash_content_pct,
    mercury_content_ppm
  ) %>%
  
  # convert "null" strings to NA
  mutate(
    mercury_content_ppm = na_if(mercury_content_ppm, "null"),
    mercury_content_ppm = as.numeric(mercury_content_ppm)
  )

#glimpse(fuel_clean)

```

Set sampling and clustering
```{r, results = 'hide'}

fuel_sample <- fuel_clean %>%
  sample_frac(0.02)

# Median imputation
median_mercury <- median(fuel_sample$mercury_content_ppm, na.rm = TRUE)

cluster_vars_q <- fuel_sample %>%
  mutate(
    mercury_content_ppm = ifelse(
      is.na(mercury_content_ppm),
      median_mercury,
      mercury_content_ppm
    )
  ) %>%
  drop_na()

# Scale variables for normalization
preProc <- preProcess(cluster_vars_q, method = c("center", "scale"))
cluster_scaled_q <- predict(preProc, cluster_vars_q)

# Cluster output
# glimpse(cluster_scaled_q)

```

Run elbow & silhouette methods to view optimal K

Fit the model and summarize
```{r, results = 'hide'}

# Elbow
wss <- map_dbl(1:10, function(k) {
  kmeans(cluster_scaled_q, centers = k, nstart = 20)$tot.withinss
})

elbow_df <- tibble(
  k = 1:10,
  wss = wss
)

# Silhouette
sil_width <- map_dbl(2:10, function(k) {
  km <- kmeans(cluster_scaled_q, centers = k, nstart = 20)
  
  ss <- silhouette(km$cluster, dist(cluster_scaled_q))
  
  mean(ss[, 3])
})

sil_df <- tibble(
  k = 2:10,
  silhouette = sil_width
)

ggplot(sil_df, aes(k, silhouette)) +
  geom_line() +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "Silhouette Method for Optimal K",
    x = "Number of clusters (K)",
    y = "Average silhouette width"
  )

# Plot elbow curve
ggplot(elbow_df, aes(x = k, y = wss)) +
  geom_line() +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "Elbow Method for Fuel Quality Clustering",
    x = "Number of clusters (K)",
    y = "Within-cluster sum of squares"
  )

k_final <- 4

final_kmeans <- kmeans(cluster_scaled_q, centers = k_final, nstart = 20)

cluster_results <- cluster_vars_q %>%
  mutate(cluster = final_kmeans$cluster)

# -------------------------
# 8. Cluster summaries
# -------------------------
cluster_results %>%
  group_by(cluster) %>%
  summarise(across(everything(), mean), .groups = "drop")

```

The elbow method would argue that K=4 is where results start to truly produce diminishing returns. Although the elbow method appears like reasonably K=4, the Silhouette analysis indicates that K=5 has a more optimized width before sharply declining at K=6, making K=5 a more stable and real clustering with the highest cohesion and separation. However, K=4 will be a little more stable with less extreme outliers. Less suggests slight under-resolved structures and more looks like over-segmentation.

View detailed profile Centroids
```{r, results = 'hide'}
km4 <- kmeans(cluster_scaled_q, centers = 4, nstart = 20)

clustered_data <- cluster_vars_q %>%
  mutate(cluster = km4$cluster)

cluster_profile <- clustered_data %>%
  group_by(cluster) %>%
  summarise(
    n = n(),
    across(
      where(is.numeric),
      list(mean = mean, sd = sd),
      .names = "{col}_{fn}"
    ),
    .groups = "drop"
  )

cluster_profile

km3 <- kmeans(cluster_scaled_q, centers = 3, nstart = 20)

table(
  K3 = km3$cluster,
  K4 = km4$cluster
)

```

Looking at the centroid group's qualities, under K=4, cluster 3 is assuming 77% of the observations and extremely low mean mmbtu. Cluster 2 and 4 are almost identical, with the key difference being their sulfur content seems to define their fuel quality grouping rather than a new fuel type, while Cluster 1 seems to be uniquely off set. The key take-away from these discoveries is that subgroup characterizations show prevalent changes that the silhouette increase alone does not distinguish. 

Going from K=3 to K=4 shown in the cross-section reveals the cluster is defined by environmental distinction rather than random assignment via high mercury content. K=5 may split an existing cluster and increase the silhouette slightly, but one cluster will be divided into two similar clusters. K=4 feels like a reasonable mid-point that yields high sulver/ash (coal), mid-grade coal, gas-like fuel, and mercury-heavy coal. Having full sets of data might prove more useful clusters for future evaluation, but there were many null values to work around mercury, moisture, and chlorine. 